Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-24 paper.qmd
Updated Figure 18
Added interp citation

## Compute hexbin parameters
num_bins_x_gau1 <- 14
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR often used to summarise the discovered clusters, and help to understand the results. The purpose of this example is to *illustrate how to use our method to help decide on an appropriate NLDR layout that accurately represents the data*. The cluster results of Human Peripheral Blood Mononuclear Cells (PBMC3k) in @chen2023 are examined. There are $2622$ single cells, with $1000$ gene expressions. Following their pre-processing, UMAP was performed using nine principal components. @fig-NLDR-variety (a) is the reproduction of the published plot, with colour indicating the clustering results overlaid on the NLDR layout. The question is whether this accurately represents the cluster structure in the data. Our method can help here, and also help to provide a better \gD{} layout, as needed.

@@

imge_error_sample_outside <- plot_digit_img(pixels_gathered_within, palette = "Grays", title_text = "(d)", hjust = 0.31)

@@

Our algorithm can also be useful in practice to decide on a reasonable NLDR layout that accurately represents the data, as described with a single-cell gene expression data, **PBMC3k**. The published results of UMAP with default (hyper-)parameter setting by @chen2023 were evaluated. The data showed three clusters, two of which were close. But the published layout showed three well-separated clusters. Therefore, by changing the (hyper-)parameter, we found a better layout with UMAP, which is more representative the data. Additionally, the model discovers non-uniform data distribution and non-linear structures within the clusters that are not visible in the UMAP layout, demonstrating the ability of our model in uncovering hidden data characteristics. Also, as explained in Appendix, with recently introduced approach, scDEED also proved the layout suggested by our layout is better than the published one.

>>>>

## Compute hexbin parameters
num_bins_x_gau1 <- 25
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR often used to summarise the discovered clusters, and help to understand the results. The purpose of this example is to *illustrate how to use our method to help decide on an appropriate NLDR layout that accurately represents the data*. The cluster results of Human Peripheral Blood Mononuclear Cells (PBMC3k) in @chen2024 are examined. There are $2622$ single cells, with $1000$ gene expressions. Following their pre-processing, UMAP was performed using nine principal components. @fig-NLDR-variety (a) is the reproduction of the published plot, with colour indicating the clustering results overlaid on the NLDR layout. The question is whether this accurately represents the cluster structure in the data. Our method can help here, and also help to provide a better \gD{} layout, as needed.

@@

imge_error_sample_outside <- plot_digit_img(pixels_gathered_outside, palette = "Grays", title_text = "(d)", hjust = 0.31)

@@

Our algorithm can also be useful in practice to decide on a reasonable NLDR layout that accurately represents the data, as described with a single-cell gene expression data, **PBMC3k**. The published results of UMAP with default (hyper-)parameter setting by @chen2024 were evaluated. The data showed three clusters, two of which were close. But the published layout showed three well-separated clusters. Therefore, by changing the (hyper-)parameter, we found a better layout with UMAP, which is more representative the data. Additionally, the model discovers non-uniform data distribution and non-linear structures within the clusters that are not visible in the UMAP layout, demonstrating the ability of our model in uncovering hidden data characteristics. Also, as explained in Appendix, with recently introduced approach, scDEED also proved the layout suggested by our layout is better than the published one.
